### 2024-02-16 misc/expo/entropy.tex
Progress in entropy

\def\eps{\epsilon}
\def\FF{{\bf F}}
\def\bar#1{\overline{#1}}
\def\hat#1{\widehat{#1}}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
\def\Bohr{\op{\rm Bohr}}
\def\Eta{\op{\bf H}}
\def\II{\op{\bf I}}
\def\dd{\op{\bf d}}
\def\given{\mathbin{|}}

@@

\maketitle{Entropy and additive combinatorics}{by}{Marcel K. Goh}{16 January 2024}

\floattext4.5 \ninebf Note.
\ninepoint This is a gentle introduction to the notion of entropy as it is used
in additive combinatorics, with a view towards understanding the new proof of the polynomial Freiman--Ruzsa
conjecture by W.~T.~Gowers, B.~Green, F.~Manners, and T.~Tao. The preliminary portion of
these notes are largely transcribed from lectures given by W.~T.~Gowers.

@@

For now, we shall take it on faith that there really exists a function on random variables satisfying
these axioms. Later on, when we prove this, we will find that actually, the axioms only define entropy
up to a multiplicative constant, so we shall add the following axiom.

\medskip
\item{f)} ({\it Normalisation.}) If $X$ is uniformly distributed on $\{0,1\}$, then
$\Eta\{X\} = \log_2(e)$.
\medskip

It is actually more common (for obvious reasons, especially in computer science)
for one to set the entropy of a uniform random variable on $\{0,1\}$
to $1$, but we shall
follow the convention of Gowers et al., since the eventual goal of these notes is to understand their proof
of the polynomial Freiman--Ruzsa conjecture.

@@

We will sometimes use the notation $X^n$ to denote the vector $(X_1,\ldots,X_n)$ where the $X_i$
are independent copies of the random variable $X$.

\edef\corxn{\the\thmcount}
\proclaim Corollary \advthm. For a discrete random variable $X$, we have $\Eta\{X^n\} = n\Eta\{X\}$.

\proof Perform induction on $n$ and use the previous proposition.\slug

In the case where $X$ is the uniform distribution on $\{0,1\}$, and $X_1,\ldots,X_n$ are independent
and distributed as $X$,
we can additionally use the normalisation axiom to conclude that
$$\Eta\{X^n\} = \Eta\{X_1,\ldots,X_n\} = \sum_{i=1}^n \Eta\{X_i\} = \log_2(e)\cdot n,$$
by induction.

A similar induction on general random variables $X_1,\ldots,X_n$ gives the following statement,
often known as the {\it chain rule}.

\parenproclaim Proposition~{\advthm} (Chain rule).
Let $X_1,\ldots,X_n$ be random variables. Then
$$\Eta\{X_1,\ldots,X_n\} = \Eta\{X_1\} + \Eta\{X_2\given X_1\}
+ \cdots+\Eta\{X_n \given X_1,\ldots,X_{n-1}\}.\noskipslug$$

@@

\edef\propunifineq{\the\thmcount}
\proclaim Proposition \advthm. Let $A\subseteq B$ with $B$ finite,
let $X$ be uniformly distributed on $A$, and let $Y$ be uniformly distributed on $B$.
Then $\Eta\{X\} \le \Eta\{Y\}$, with equality if and only if $A = B$.

\proof By the extensibility axiom, $\Eta\{X\}$ is not affected if we regard $X$ as a function
taking values in $B$. Then by the maximisation axiom, $\Eta\{X\}\le \Eta\{Y\}$, since $Y$ is
uniform on $B$. If $A = B$, then it is clear that $\Eta\{X\} = \Eta\{Y\}$, since $X$ and $Y$
are the same random variable.
On the other hand, say $|A| = m$ and $|B| = n$ with $m<n$. If $m=1$, then by the previous proposition
we have $\Eta\{X\} = 0$, and by normalisation and invariance, $\Eta\{Y\} = \log_2(e)$.\slug

If $Y$ is a random variable such that $Y = f(X)$ for some random variable $Y$ and some function $f$,
then we say that $Y$ is {\it determined by} $X$ or $X$ {\it determines} $Y$.
We want to show that $\Eta\{Y\}\le \Eta\{X\}$, which reflects the idea that we get more information
from $X$ than from $Y$. This, rather annoyingly, seems to require two steps.

@@

\proof There is a a bijection between values $x$ taken
by $Y$ and values $\bigl(x,f(x)\bigr)$ taken by $(X,Y)$, so we have
$$\Eta\{X\} = \Eta\{X,Y\} = \Eta\{Y\} + \Eta\{X\given Y\}$$
by additivity.\slug

We are now done if we can show that entropy is nonnegative.
This can also be derived straight from the axioms. The following proof is due to S.~Eberhard.

\proclaim Proposition \advthm. Let $X$ be a discrete random variable taking values in a finite
set $A$. Then $\Eta\{X\} \ge 0$.

@@

Since $Z$ is determined by $Y$, we have $\Eta\{Y\} = \Eta\{Z\} + \Eta\{Y\given Z\}$ by the previous lemma.
Furthermore, for every $a\in A$, the conditional entropy $\Eta\{Y \given Z = a\}$ is uniformly
distributed on a set of size at most $n$, whereas is uniformly distributed on $n$, so by
Proposition~{\propunifineq}, we have $\pr\{Y \given Z = a\} \le \Eta\{Y\}$. This implies
that $\Eta\{Y\} \ge \Eta\{Y\given Z\}$, which in turn gives us $\Eta\{Z\}\ge 0$.\slug

\proclaim Corollary \advthm. If $Y = f(X)$ then $\Eta\{X\} \ge \Eta\{Y\}$.\slug

Now we show that a random variable has zero entropy if and only if it is constant.
This reflects the idea that the variables from which we get no information are those which take
the same value no matter what.

@@

\medskip\boldlabel Mutual information.
For random variables $X$ and $Y$, the {\it mutual information} $\II\{X : Y\}$ is defined
by the equivalent formulas
$$\eqalign{
\II\{ X : Y\} &= \Eta\{X\} + \Eta\{Y\} - \Eta\{X,Y\} \cr
&= \Eta\{X\} - \Eta\{X\given Y\} \cr
&= \Eta\{Y\} - \Eta\{Y\given X\}. \cr
}$$
It measures, roughly speaking, how much information one can get from one variable
by looking at the other one. From the formula it is clear that $\II\{X : Y\}  = 0$
if and only if $X$ and $Y$ are independent, and we also have the
inequality $\Eta\{X \given Y\} \le \Eta\{X\}$; that is, {\sl conditioning cannot increase the
entropy of a random variable}. Given a triple $(X,Y,Z)$ of random variables,
we can apply this with $(X\given Z = z)$ to obtain the inequality
$$ \Eta\{ X\given Y,Z\} \le \Eta\{X\given Z\},$$
which is called {\it submodularity}. An equivalent statement is
$$\Eta\{X,Y,Z\} + \Eta\{Z\} \le \Eta\{X,Z\} + \Eta\{Y,Z\}.$$
If $Z$ takes values in a set $C$, the {\it conditional mutual information} is defined by
$$\eqalign{
\II\{ X : Y \given Z\} &= \sum_{z\in C} p_Z(z) \II\bigl\{ (X\given Z = z) : (Y\given Z = z)\bigr\} \cr
&= \Eta\{ X\given Z\} + \Eta\{Y\given Z\} - \Eta\{ X, Y \given Z \} \cr
&= \Eta\{ X, Z\} -2\Eta\{ Z\} + \Eta\{Y,Z\} - \Eta\{ X, Y ,Z \} + \Eta\{Z\}\cr
&= \Eta\{ X, Z\} + \Eta\{Y,Z\} - \Eta\{ X, Y ,Z \} - \Eta\{Z\},\cr
}$$
so we see that submodularity of entropy is equivalent to the statement that $\II\{X : Y\given Z\} \ge 0$.
Analogously to the unconditional case, we have equality if and only if $X$ and $Y$ are
independent when conditioned on $Z$.

@@

\edef\maxsumsetbound{\the\thmcount}
\proclaim Proposition \advthm. Let $X$ and $Y$ be random variables each taking finitely
many values in an abelian group $G$.
We have
$$\max\bigl( \Eta\{X\}, \Eta\{Y\} \bigr) - \II\{X : Y\} \le \Eta\{ X \pm Y\}.$$
Furthermore, for any random variable $Z$, we have the conditional version
$$\max\bigl( \Eta\{X\given Z\}, \Eta\{Y\given Z\} \bigr) - \II\{X : Y\given Z\} \le \Eta\{ X \pm Y\given Z\}$$
of the same statement.

@@

\medskip\boldlabel Entropic Ruzsa distance.
In additive combinatorics, whenever we have two finite subsets $A$ and $B$
of the same abelian group, we can compute the Ruzsa distance
$$d(A,B) = \log {|A-B|\over \sqrt{|A|\cdot|B|}}$$
between them. (This satisfies all the axioms of a metric except the one requiring $d(A,A) = 0$ for all sets
$A$.)

@@

@@

\parenproclaim Theorem A (Pl\"unnecke--Ruzsa inequality). Let $A$ and $B$ be finite subsets of an
abelian group and suppose that $|A+B|\le K|A|$ for some constant $K$. Then for any integers
$r,s\ge 0$, not both zero, we have $|rB-sB| \le K^{r+s}|A|$.\slug

@@

% \medskip\boldlabel The formula for entropy.
% Now we finally give the formula for the entropy of a random variable. For a discrete random
% variable $X$, we let $p_X : S\to [0,1]$ denote the probability mass function of $X$, given by
% $p_X(x) = \pr\{X = x\}$ for all $x\in S$. Unless subscripted, $\log$ always denotes the natural logarithm.
% 
% \proclaim Proposition \advthm. Let $X$ be a random variable taking values in a finite set $S$.
% Then $\Eta\{X\}$ satisfies axioms (a) through (e) as well as the additional normalisation axiom (f)
% if and only if
% $$\Eta\{X\} = \ex\bigl\{ 1/\log p_X(X) \bigr\} = \sum_{x\in S} p_X(x) \log {1\over p_X(s)},$$
% where we adopt the convention that $0\log(1/0) = 0$.
% 
% \proof [TODO: Prove this.]\slug
% 
% Equipped now with a formula for the entropy $\Eta\{X\}$ of a random variable $X$, we can prove further useful
% facts. We saw in the previous section that the entropy of a uniform random variable whose range is $\{0,1\}^n$
% is $\log_2(e)\cdot n = \log(2^n)$, and we know, by the maximisation axiom, that this is the largest
% the entropy of any $X$ taking values in $\{0,1\}^n$ can be. The following proposition generalises
% this notion.
% 
% \proclaim Proposition \advthm. For any random variable $X$ taking values in a finite set $S$,
% we have
% $$\Eta\{X\} \le \log |S|.$$
% 
% \proof By Jensen's inequality, we have
% $$\eqalign{
% \Eta\{X\} &= \ex\bigl\{ \log(1/p_X(X)\bigr\} \cr
% &\le \log \ex\bigl\{ 1/p_X(X)\bigr\} \cr
% &= \log \sum_{x\in S} \pr\{X = x\} {1\over \pr\{X = x\}} \cr
% &= \log |S|.\noskipslug \cr
% }$$

>>>>

\def\eps{\epsilon}
\def\FF{{\bf F}}
\def\bar#1{\overline{#1}}
\def\hat#1{\widehat{#1}}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
\def\Bohr{\op{\rm Bohr}}
\def\Eta{\op{\bf H}}
\def\Etaseven{\op{\sevenbf H}}
\def\II{\op{\bf I}}
\def\dd{\op{\bf d}}
\def\ee{\op{\bf e}}
\def\given{\mathbin{|}}
\def\argmax{\limitop{arg$\;$max}}
\def\advthm{\the\sectcount.\the\thmcount\global\advance \thmcount by 1}
\def\advsect{\global\advance\sectcount by 1\section\the\sectcount\global\thmcount=1. }
\sectcount=0

@@

\maketitle{Entropy and additive combinatorics}{by}{Marcel K. Goh}{15 February 2024}

\floattext4.5 \ninebf Abstract.
\ninepoint These expository notes give a gentle introduction to the notion of entropy as it is used
in additive combinatorics, with a view towards understanding the proof of the polynomial Freiman--Ruzsa
conjecture by W.~T.~Gowers, B.~Green, F.~Manners, and T.~Tao. Much of the first section has been
transcribed from lectures given by W.~T.~Gowers.

@@

We shall take it on faith that there really exists a function on random variables satisfying
these axioms.
In fact, the axioms only define entropy
up to a multiplicative constant, so we shall add the following axiom.
(It is very possible you have met this function somewhere on your travels, but you will not find it anywhere in
these notes.)

\medskip
\item{f)} ({\it Normalisation.}) If $X$ is uniformly distributed on $\{0,1\}$, then
$\Eta\{X\} = 1$.
\medskip

@@

We will sometimes use the notation $X^n$ to denote the vector $(X_1,\ldots,X_n)$ where the $X_i$
are independent copies of the random variable $X$. We have the following three corollaries, which
are each proved by induction. The second also requires the normalisation axiom,
and the third is often known as the {\it chain rule}.

\edef\corxn{\the\sectcount.\the\thmcount}
\proclaim Corollary \advthm. We have $\Eta\{X^n\} = n\Eta\{X\}$.\slug

\edef\cortwon{\the\sectcount.\the\thmcount}
\proclaim Corollary \advthm. If $X$ is uniformly distributed on a set of size $2^n$, then
$$\Eta\{X\} = n.\noskipslug$$

\parenproclaim Corollary~{\advthm} (Chain rule).
Let $X_1,\ldots,X_n$ be random variables. Then
$$\Eta\{X_1,\ldots,X_n\} = \Eta\{X_1\} + \Eta\{X_2\given X_1\}
+ \cdots+\Eta\{X_n \given X_1,\ldots,X_{n-1}\}.\noskipslug$$

@@

\edef\propunifineq{\the\sectcount.\the\thmcount}
\proclaim Proposition \advthm. Let $A\subseteq B$ with $B$ finite,
let $X$ be uniformly distributed on $A$, and let $Y$ be uniformly distributed on $B$.
Then $\Eta\{X\} \le \Eta\{Y\}$, with equality if and only if $A = B$.

\proof By the extensibility axiom, $\Eta\{X\}$ is not affected if we regard $X$ as a function
taking values in $B$. Then by the maximisation axiom, $\Eta\{X\}\le \Eta\{Y\}$, since $Y$ is
uniform on $B$.

If $A = B$, then it is clear that $\Eta\{X\} = \Eta\{Y\}$, since $X$ and $Y$
are the same random variable.

On the other hand, say $|A| = m$ and $|B| = n$ with $m<n$. If $m=1$, then by the previous proposition
we have $\Eta\{X\} = 0$, and by normalisation and invariance, $\Eta\{Y\} = 1$.
When $m\ge 2$, pick $k$ such that $m^k \le n^{k-1}$, so that $|A^k| \le |B^{k-1}|$.
Then by Corollary~{\corxn} and the inequality we showed in the first paragraph of this proof, we have
$$n\Eta\{X\} = \Eta\{X^n\} \le \Eta\{X^{n-1}\} = (n-1)\Eta\{Y\},$$
whence $\Eta\{X\} < \Eta\{Y\}$.\slug

The link between entropy and additive combinatorics is rather a fine one. It is based on the following
observation. (In this and the rest of the notes, $\lg$ denotes the base-$2$ logarithm. This base can
be changed by modifying the normalisation axiom.)

\edef\proplogeq{\the\sectcount.\the\thmcount}
\proclaim Proposition \advthm. Let $X$ be a uniform random variable on a finite set $A$.
Then
$$\Eta\{X\} = \lg |A|.$$

\proof For any positive integer $n$ we can let
$X^n$ denote a tuple of independent copies of $X$; Corollary~{\corxn} tells us $\Eta\{X^n\} = n\Eta\{X\}$.
Let $m$ be such that $2^m \le |A|^n \le 2^{m+1}$ so that
$${m\over n} \le \lg |A| \le {(m+1)\over n}.$$
Let $Y$ be uniform on a set of size $2^m$,
and let $Z$ be uniform on a set of size $2^{m+1}$. Then by Corollary~{\cortwon} we have
$\Eta\{Y\} = m $ and $\Eta\{Z\} = (m+1)$. Then by Proposition~{\propunifineq}
we have
$${m\over n} \le \Eta\{X\} \le {(m+1)\over n}.$$
In other words, $\Eta\{X\}$ satisfies the same bounds as $\lg |A|$. Taking $n$ large, we can make
these bounds arbitrarily tight, proving the claim.\slug

The maximisation axiom gives the following corollary.

\edef\corlogineq{\the\sectcount.\the\thmcount}
\proclaim Corollary \advthm. Let $X$ be a random variable supported on a finite set $A$.
Then
$$\Eta\{X\} \le \lg |A|.\noskipslug$$

Hence the entropy $\Eta\{X\}$ is at most the exponential of the size of its support. As we will see,
simply replacing (logarithms of) cardinalities with entropies, we at useful ``entropic analogues''
of combinatorial statements. But first, we need more lemmas.

If $Y$ is a random variable such that $Y = f(X)$ for some random variable $Y$ and some function $f$,
then we say that $Y$ is {\it determined by} $X$ or $X$ {\it determines} $Y$.
We want to show that $\Eta\{Y\}\le \Eta\{X\}$, which reflects the idea that we get more information
from $X$ than from $Y$. This, rather annoyingly, seems to require a couple of steps.

@@

\proof There is a a bijection between values $x$ taken
by $X$ and values $\bigl(x,f(x)\bigr)$ taken by $(X,Y)$, so we have
$$\Eta\{X\} = \Eta\{X,Y\} = \Eta\{Y\} + \Eta\{X\given Y\}$$
by invariance and additivity.\slug

We are now done if we can show that entropy is nonnegative.
This is a corollary of the following lemma. The following proof is a modification of
one due to S.~Eberhard.

\edef\lemmaxprob{\the\sectcount.\the\thmcount}
\proclaim Proposition \advthm. Let $X$ be a discrete random variable supported on a set $A$ and let
$$a^* = \argmax_{a\in A} \pr\{X = a\}.$$
Then
$$\pr\{X = a^*\} \ge 2^{-\Etaseven\{X\}}.$$

@@

For every $a\in A$, the conditional entropy $\Eta\{Y \given Z = a\}$ is uniformly
distributed on a set of size $|E_a|$. From our choice of $a^*$
we have $|E_{a^*}| \ge |E_a|$ for all $a\in A$. Hence by Proposition~{\proplogeq}, we have
$$\Eta\{Y\given Z\} = \sum_{a\in A} \pr\{X = a\} \Eta\{Y\given X = a\}
\sum_{a\in A} \pr\{X = a\} \log |E_a| \le \log |E_{a^*}|.$$
Since $Z$ is determined by $Y$, we have $\Eta\{Y\} = \Eta\{Z\} + \Eta\{Y\given Z\}$ by the previous lemma.
So by another invocation of Proposition~{\proplogeq}, we have
$$\eqalign{
\Eta\{Z\} &= \Eta\{Y\} - \Eta\{Y\given Z\} \cr
&\ge \log n - \log |E_{a^*}| \cr
&\ge \log \biggl({n\over |E_{a^*}|}\biggr) \cr
&= \log\biggl({1\over \pr\{X = a^*\}}\biggr), \cr
}$$
and hence $2^{-\Etaseven\{X\}} = 2^{-\Etaseven\{Z\}} \le \pr\{X = a^*\}$.

The general case follows from the continuity axiom.\slug

This proof came dangerously close to deriving the formula for entropy, but we will not need any such formula,
so we will refrain from mentioning it.
From the fact that $\pr\{X = a^*\} \le 1$ we can immediately conclude that entropy is nonnegative.

\edef\corentropynonneg{\the\sectcount.\the\thmcount}
\proclaim Corollary \advthm. Let $X$ be a discrete random variable taking values in a finite
set $A$. Then $\Eta\{X\} \ge 0$.\slug

This observation completes the proof that a random variable has a smaller entropy than one by which
it is determined.

\edef\cordetermines{\the\sectcount.\the\thmcount}
\proclaim Corollary \advthm. If $Y = f(X)$ then $\Eta\{X\} \ge \Eta\{Y\}$.\slug

Next we show that a random variable has zero entropy if and only if it is constant.
This reflects the idea that the variables from which we get no information are those which take
the same value no matter what.

@@

\medskip\boldlabel Mutual information.
For random variables $X$ and $Y$, the {\it mutual information} $\II\{X : Y\}$ is defined
by the equivalent formulas
$$\eqalign{
\II\{ X : Y\} &= \Eta\{X\} + \Eta\{Y\} - \Eta\{X,Y\} \cr
&= \Eta\{X\} - \Eta\{X\given Y\} \cr
&= \Eta\{Y\} - \Eta\{Y\given X\}. \cr
}$$
It measures, roughly speaking, how much information one can get from one variable
by looking at the other one. From the formula it is clear that $\II\{X : Y\}  = 0$
if $X$ and $Y$ are independent. In general, we still have the inequality $\II\{X : Y\} \ge 0$,
which a corollary of the following lemma, which expresses the intuitive fact that conditioning
cannot increase the entropy of a random variable.
The proof, which is due to C.~West, uses an argument similar to the one
we used to prove Lemma~{\lemmaxprob}.

\proclaim Proposition {\advthm}. Let $X$ and $Y$ be discrete random variables. Then
$$\Eta\{X\given Y\} \le \Eta\{X\}.$$

\proof Let $A$ be the support of $X$ and $B$ be the support of $Y$.
First we consider the case that $X$ is uniform on $A$ (so $A$ is finite). Then by the definition
of conditional entropy,
$$\Eta\{X\given Y\} = \sum_{b\in B} \pr\{Y = b\} \Eta\{X \given Y= b\}.$$
But for each $b$, the random variable $(X\given Y = b)$ takes values in $A$, so its entropy is
bounded above by $\Eta\{X\}$ by the maximisation axiom. Hence $\Eta\{X\given Y\} \le \Eta\{X\}$.

Next, suppose that $A$ and $B$ are both finite and suppose further that $\pr\{Y = b\}$
is rational for all $b$. Then there is an integer $n$ and integers $\{m_b\}_{b\in B}$
such that $\pr\{Y= b\} = m_b/n$ for all $b\in B$. Now partition $[n]$ into sets $\{E_b\}_{b\in B}$,
where $|E_b| = m_b$ for all $b\in B$. We define a random variable $Z$ by sampling uniformly at
random from $E_b$ if $Y = b$, and doing so independently of $(X\given Y=b)$. The result is
a random variable $Z$ that is uniform on $[n]$ and which is independent of $X\given Y$.
Furthermore, since $Z$ determines $Y$, we have $\Eta\{Z\} = \Eta\{Y,Z\}$ by the invariance
axiom. Hence
$$\eqalign{
\Eta\{X\given Y\} &= \Eta\{X\given Y, Z\} \cr
&= \Eta\{X,Y,Z\} - \Eta\{Y,Z\} \cr
&= \Eta\{X,Z\} - \Eta\{Z\} \cr
&= \Eta\{X\given Z\} \cr
&\le \Eta\{X\},\cr
}$$
where the inequality on the last line follows from the previous paragraph.

The general case follows from the continuity axiom and the fact that any discrete random
variable can be approximated by discrete random variables with finite support and rational
probabilities.\slug

By the additivity axiom, the previous proposition is equivalent to
$$\Eta\{X,Y\} \le \Eta\{X\} + \Eta\{Y\}.$$
and we shall use this to prove the following submodularity inequality.

\parenproclaim Proposition {\advthm} (Submodularity).
Suppose $X$, $Y$, $Z$, and $W$ are random variables such that $(Z,W)$ determines
$X$, $Z$ determines $Y$, and $W$ also determines $Y$. Then
$$\Eta\{X\} + \Eta\{Y\} \le \Eta\{Z\} + \Eta\{W\}.$$

\proof The hypotheses give the three inequalities
$$\Eta\{X\} \le \Eta\{Z,W\}, \quad \Eta\{Y\}\le \Eta\{Z\},\quad\hbox{and}\quad \Eta\{Y\}\le \Eta\{W\}.$$
From this we see that
$$2\Eta\{X\} + 2\Eta\{Y\} \le 2\Eta\{Z,W\} + \Eta\{Z\} + \Eta\{Y\}.$$
But then since conditioning does not increase entropy, we have
$$2\Eta\{X\} + 2\Eta\{Y\} \le 2\Eta\{Z\} + 2\Eta\{Y\},$$
whence dividing both sides by $2$ completes the proof.\slug

The submodularity inequality is often stated in terms of a triple of random variables in terms of
the {\it conditional mutual information}, which is defined by
$$\II\{ X : Y \given Z\} = \sum_{z\in C} p_Z(z) \II\bigl\{ (X\given Z = z) : (Y\given Z = z)\bigr\}.$$

\proclaim Proposition \advthm. Let $X$, $Y$, and $Z$ be discrete random variables. Then
$$\Eta\{X,Y,Z\} + \Eta\{Z\} \le \Eta\{X,Z\} + \Eta\{Y,Z\},$$
which is equivalent to
$$\II\{X : Y\given Z\} \ge 0.$$

\proof It is clear from the additivity axiom that both sides in the first
inequality are equal to $\Eta\{X\} + \Eta\{Y\} + 2\Eta\{Z\}$
if and only if $X$ and $Y$ are independent conditional on $Z$.

To prove the first inequality, note that $(X,Y,Z)$ is jointly determined by $(X,Z)$ and $(Y,Z)$,
and $Z$ is determined by both $(X,Z)$ and $(Y,Z)$ separately, then apply the previous proposition.
Now by the definition of conditional mutual information,
$$\eqalign{
\II\{ X : Y \given Z\} &= \sum_{z\in C} p_Z(z) \II\bigl\{ (X\given Z = z) : (Y\given Z = z)\bigr\} \cr
&= \Eta\{ X\given Z\} + \Eta\{Y\given Z\} - \Eta\{ X, Y \given Z \} \cr
&= \Eta\{ X, Z\} -2\Eta\{ Z\} + \Eta\{Y,Z\} - \Eta\{ X, Y ,Z \} + \Eta\{Z\}\cr
&= \Eta\{ X, Z\} + \Eta\{Y,Z\} - \Eta\{ X, Y ,Z \} - \Eta\{Z\},\cr
}$$
and this proves that the first statement is equivalent to the second.\slug

@@

\edef\maxsumsetbound{\the\sectcount.\the\thmcount}
\proclaim Proposition \advthm. Let $X$ and $Y$ be random variables each taking finitely
many values in an abelian group $G$.
We have
$$\max\bigl( \Eta\{X\}, \Eta\{Y\} \bigr) - \II\{X : Y\} \le \Eta\{ X \pm Y\}.$$
Furthermore, for any random variable $Z$, we have the conditional version
$$\max\bigl( \Eta\{X\given Z\}, \Eta\{Y\given Z\} \bigr) - \II\{X : Y\given Z\} \le \Eta\{ X \pm Y\given Z\}$$
of the same statement.

@@

\medskip\boldlabel Entropic Ruzsa distance.
In additive combinatorics, whenever we have two finite subsets $A$ and $B$
of the same abelian group, we can compute the Ruzsa distance
$$d(A,B) = \lg {|A-B|\over \sqrt{|A|\cdot|B|}}$$
between them. (This satisfies all the axioms of a metric except the one requiring $d(A,A) = 0$ for all sets
$A$.)

@@

\medskip\boldlabel The sum-difference inequality.
The Ruzsa triangle inequality bounds the size of a difference set by passing through a different subset. There
is another inequality that relates the size of a sumset with the size of a difference set.
If $A$ and $B$ are nonempty finite subsets of an abelian group $G$, then
$$|A+B| \le {|A-B|^3\over |A||B|}.$$
If we replace cardinalities by exponentials of entropies, then we obtain the statement of the following
proposition.

\proclaim Proposition \advthm.
Let $X$ and $Y$ be independent random variables taking values in the same abelian group. Then
$$\Eta\{X+Y\} \le 3\Eta\{X-Y\} - \Eta\{X\} - \Eta\{Y\}.$$

Before we proceed to the proof, we establish the following definition, which will be needed later in these
notes as well. Let $X$ and $Y$ be random variables (not necessarily independent). We say that
$X_1$ and $Y_1$ are {\it conditionally independent trials of $X$ and $Y$ relative to $Z$} if for all
$z$ in the range
of $Z$, the random variables distributed as $(X_1\given Z=z)$ and $(Y_1\given Z=z)$ are independent,
$(X_1\given Z=z)$ has the same distribution as $(X\given Z=z)$,
and similarly for $Y_1$ and $Y$.
In particular, if $X=Y$ and $X_1$ and $X_2$ are conditionally independent trials of $X$ relative
to $Z$, we have
$$\Eta\{X_1,X_2\given Z\} = \Eta\{X_1\given Z\} + \Eta\{X_2\given Z\} = 2\Eta\{X\given Z\},$$
by additivity and independence. From this we obtain
\edef\eqcondindep{\the\eqcount}
$$\Eta\{X_1,X_2,Z\} = 2\Eta\{X\given Z\} + \Eta\{Z\} = 2\Eta\{X,Z\} - \Eta\{Z\}.\adveq$$
It is also important to observe that $(X_1,Z)$ and $(X_2,Z)$ both have the same distributions
as $(X,Z)$.

\proof Let $(X_1,Y_1)$ and $(X_2,Y_2)$ be conditionally independent trials of $(X,Y)$ relative
to $X-Y$. Since $(X,Y)$ determines $X-Y$, we have $X_1-Y_1 = =X-Y= X_2-Y_2$. Let $(X_3,Y_3)$ be another
trial of $(X,Y)$ independent of $(X_1, X_2, Y_1, Y_2)$. Then
$$X_3 + Y_3 = X_3 + Y_3 + X_1 - Y_1 - X_2 + Y_2 = (X_3-Y_2) - (X_1-Y_3) + X_2 + Y_1,$$
so $(X_3-Y_2, X_1-Y_3, X_2, Y_1)$ and $(X_3, Y_3)$ each determine $X_3+ Y_3$. On the other hand,
$(X_3-Y_2, X_1-Y_3, X_2, Y_1)$ and $(X_3,Y_3)$ together determine the sextuple $(X_1,X_2,X_3,Y_1,Y_2,Y_3)$,
so by the submodularity inequality, we have
\edef\eqsumdiff{\the\eqcount}
$$\eqalign{
\Eta\{X_1,X_2,X_3,Y_1,Y_2,&Y_3\} + \Eta\{X_3+Y_3\} \cr
&\le \Eta\{X_3-Y_2, X_1-Y_3, X_2, Y_1\} + \Eta\{X_3, Y_3\}.\cr
}\adveq$$
We have
$$\Eta\{X_3,Y_3\} = \Eta\{X,Y\} = \Eta\{X\} + \Eta\{Y\}$$
by independence of
$X$ and $Y$, and
since $(X_3,Y_3)$ and $(X_1,X_2,Y_1,Y_2)$ are independent, we have
$$\eqalign{
\Eta\{X_1,X_2,X_3,Y_1,Y_2,Y_3\} &= \Eta\{X_1,X_2,Y_1,Y_2\} + \Eta\{X_3,Y_3\} \cr
&= \Eta\{X_1, Y_1,X_2,Y_2,X-Y\} + \Eta\{X\} + \Eta\{Y\} \cr
&= 2\Eta\{X,Y,X-Y\} - \Eta\{X-Y\} + \Eta\{X\} + \Eta\{Y\} \cr
&= 2\Eta\{X,Y\} - \Eta\{X-Y\} + \Eta\{X\} + \Eta\{Y\}\cr
&= 3\Eta\{X\} + 3\Eta\{Y\} - \Eta\{X-Y\},\cr
}$$
where in the third line we applied~\refeq{\eqcondindep}. On the other hand,
$$\Eta\{X_3 + Y_3\} = \Eta\{X+Y\}$$
and
$$\eqalign{
\Eta\{X_3-Y_2, &X_1-Y_3, X_2, Y_1\} \cr
&\le \Eta\{X_3-Y_2\} + \Eta\{X_1-Y_3\} + \Eta\{X_2\} + \Eta\{Y_1\} \cr
&= 2\Eta\{X-Y\} + \Eta\{X\} + \Eta\{Y\}.
}$$
Substituting everything into~\refeq{\eqsumdiff} yields
$$\eqalign{
3\Eta\{X\} + 3\Eta\{Y\} - &\Eta\{X-Y\} + \Eta\{X+Y\} \cr
&\le 2\Eta\{X-Y\} + 2\Eta\{X\} + 2\Eta\{Y\},
}$$
and the desired inequality follows upon rearrangement of terms.\slug

The entropic sum-difference inequality can also be stated in terms of entropic Ruzsa distances;
independence is not necessary here because independent trials are baked into the definition
of entropic Ruzsa distance.

\proclaim Corollary \advthm. Let $X$ and $Y$ be discrete random variables taking values in the same
abelian group. Then
$$\dd\{X, -Y\} \le 3\dd\{X,Y\}.\noskipslug$$

@@

\parenproclaim Theorem {\advthm} (Pl\"unnecke--Ruzsa inequality). Let $A$ and $B$ be finite subsets of an
abelian group and suppose that $|A+B|\le K|A|$ for some constant $K$. Then for any integers
$r,s\ge 0$, not both zero, we have $|rB-sB| \le K^{r+s}|A|$.\slug

@@

\advsect The Balog--Szemer\'edi--Gowers theorem

If $A$ and $B$ are subsets of the same abelian group such that $|A+B|$ is much smaller than $|A|\cdot|B|$,
then it stands to reason that there must be a lot of redundancy in $A+B$; that is, many elements of $A+B$ can
be expressed as $a+b$ in lots of different ways. To capture this notion, we can define the
{\it additive energy} between two sets $A$ and $B$ to be
$$E(A,B) = \bigl| \bigl\{ (a,a',b,b') \in A\times A\times B\times B : a + b = a'+b'\bigr\}\bigr\}.$$

We now define an entropic version of $E(A,B)$. Let $X$ and $Y$ be discrete random variables taking values
in the same
abelian group. Let $(X_1, Y_1)$ and $(X_2, Y_2)$ be conditionally independent trials of $(X,Y)$ relative to
$X+Y$. These  $X_1 + Y_1 = X_2 + Y_2 = A+B$.
The {\it entropic additive energy} between $X$ and $Y$ is
$$ \ee\{X,Y\} = \Eta\{X_1, Y_1, X_2, Y_2\}.$$
This definition makes clear the analogy between this value and the additive energy of sets, but
by conditional independence and the fact that $(X_1, Y_1, X_2, Y_2)$ determines $X_1+Y_1 = X+Y$,
we can rewrite
$$\eqalign{
\ee\{X,Y\} &= \Eta\{X_1, Y_1, X_2, Y_2, X+Y\} \cr
&= 2\Eta\{X,Y,X+Y\} - \Eta\{X+Y\} \cr
&= 2\Eta\{X,Y\} - \Eta\{X+Y\},\cr
}$$
where in the second equality we applied~\refeq{\eqcondindep}. This formula is something we will use
often, as it makes no direct mention of the variables $(X_1, Y_1)$ and $(X_2,Y_2)$.

Quantifying the idea that small sumset must imply large additive energy, we have the following
proposition.

\edef\propinversebalog{\the\sectcount.\the\thmcount}
\proclaim Proposition \advthm.
Let $A$ and $B$ be finite subsets of an abelian group.
If $|A+B|\le K|A|^{1/2} |B|^{1/2}$ for some constant $K$, then we have
$$E(A,B) \ge {1\over K} |A|^{3/2} |B|^{3/2}.\noskipslug$$

Somewhat surprisingly, if we convert
these statements into their entropic analogues in the na\"ive way, as we've been doing,
the implication goes the other way!
However, we have a weak equivalence (with worse constants in one direction)
under the further assumption that the random variables in
question are not too dependent.
